Supporting information

### 2023-10-31 paper/main.qmd
Update paper

![](./figures/methionine-reactions.png)

@@

![](./figures/missing.png)

@@

![](./figures/decomposition.png)

@@

Maud is a command line application implementing Bayesian inference for
a wide range of realistic kinetic models. Maud is written in Python
[@vanrossumPythonReferenceManual2009], designed for use on Windows, macOS
and Linux, registered on the Python Package Index as `maud-metabolic-models`,
documented at <https://maud-metabolic-models.readthedocs.io> and actively
developed and maintained at <https://github.com/biosustain/Maud/>.

To use Maud, a user must first collate appropriate input information, represent
it in files with Maud’s required formats (see section (REFERENCE)below).  Maud’s
command line interface provides commands for inference, simulation and making
out-of-sample predictions. Results are stored in files, using a structured,
interoperable format.

## Input format

Maud inputs are structured directories, somewhat inspired by the PEtab
format [@SchmiesterSch2021]. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.

We chose to implement a custom input format despite the existence of standard
formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features of
SBML and PEtab that make them structurally unsuitable in this context. Our
requirements for an input format included that it be mathematics-free, so that
all mathematical details are encapsulated in source code, and that it has a
detailed, verifiable structure. These requirements made toml more attractive
than SBML: toml is easier for humans to read and edit and can straightforwardly
be validated using tools like Pydantic. Further, an SBML representation of our
desired input would not contain differential equations.  It would therefore not
be interoperable with most SBML targeting software, which typically assumes that
differential equations are available and does not know about Maud’s structure.

@@

Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
to implement a command line interface. The command line interface loads input
files as Python dictionaries, which are parsed using the Python library `toml`
[@pearsonTomlPythonLibrary2020] and then validated and converted into structured
`MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022]. Maud’s
statistical model is implemented in the probabilistic programming language Stan
[@carpenterStanProbabilisticProgramming2017] and accessed using the interface
cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior sampling, Maud
uses the `MaudInput` to create a Stan input json file and obtain configuration
information for cmdstanpy, which it uses to (if necessary) generate a
model executable file and then trigger posterior sampling using adaptive
Hamiltonian Monte Carlo. When sampling is complete, Maud converts to the
output into the standard format `InferenceData` using the Python library arviz
[@kumarArviZUnifiedLibrary2019] and saves it as a json file, along with some
information for debugging.

Two details of Maud’s implementation are important to highlight: the method of
posterior sampling and how Maud solves the steady state problem \eqref{eq-3}.

### Posterior sampling

Although integrals of the joint probability model for kinetic models
are typically analytically intractable, they can be approximated
numerically using Markov Chain Monte Carlo (MCMC) and other methods.
Maud uses MCMC primarily because there exist many methods for verifying
that MCMC samples really do approximate the target probability
distribution: see @vehtariRankNormalizationFoldingLocalization2021 and
@taltsValidatingBayesianInference2018 for discussion of this point. In
addition, there are several examples of successful Bayesian kinetic modelling
projects using MCMC including @st.johnBayesianInferenceMetabolic2018 and
@xingModelingKineticsLargescale2010.

### Solving the steady state problem

In order to implement adaptive Hamiltonian Monte Carlo, Maud must
repeatedly solve the steady state problem \eqref{eq-3} and find its
gradients with respect to all parameters. To achieve this, Maud uses a
hybrid method involving two numerical solvers from the SUNDIALS suite
[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS. These solvers are
accessed via their interface from Stan. The hybrid method follows that proposed
by @margossianComputingSteadyStates2018, and involves numerically evolving
the ODE system for a short period of time, then using the difference between
the evolved and starting concentrations as the target for a numerical algebra
solver.

>>>>

![](../figures/methionine-reactions.png)

@@

![](../figures/missing.png)

@@

![](../figures/decomposition.png)

@@

Maud is a command line application implementing Bayesian inference
for a wide range of realistic kinetic models. Maud is written in
Python [@vanrossumPythonReferenceManual2009], designed for use
on Windows, macOS and Linux, is `pip` installable from the Python
Package Index as `maud-metabolic-models`, documented at <https://maud-metabolic-models.readthedocs.io> 
and actively developed and maintained at <https://github.com/biosustain/Maud/>.

To use Maud, a user must first collate appropriate input information, represent
it in files with Maud’s required formats (see supporting information section on
input format).  Maud’s command line interface provides commands for inference
via a range of algorithms including adaptive Hamiltonian Monte Carlo, as well as
commands for simulation and making out-of-sample predictions. Results are stored
in files, using a structured, interoperable format.

As well as parameter values, Maud also performs inference for derived quantities
including the components of the regulatory decomposition described below in
equation \eqref{eq-decomposition}, log likelihoods, simulated measurement values
and metabolic control analysis coefficients [@kacser_control_1973].

In the rest of this section we describe Maud's kinetic model at high level and
discuss Maud's statistical model, implementation and how it solves the crucial
steady state problem. Further details about Maud's kinetic model can be found in
the supporting information.

@@

Maud uses the Python library click [@clickdevelopersClickPythonComposable2022]
to implement a command line interface. The command line interface loads
input files as Python dictionaries, which are parsed using the Python library
`toml` [@pearsonTomlPythonLibrary2020] and then validated and converted into
structured `MaudInput` objects using Pydantic [@pydanticdevelopersPydantic2022].
Maud’s statistical model is implemented in the probabilistic programming
language Stan [@carpenterStanProbabilisticProgramming2017] and accessed using
the interface cmdstanpy [@standevelopmentteamCmdStanPy2022]. For posterior
sampling, Maud uses the `MaudInput` to create an input file for Stan and
then trigger posterior sampling using adaptive Hamiltonian Monte Carlo. See
@betancourtConceptualIntroductionHamiltonian2018 for more about this algorithm.

When sampling is complete, Maud converts to the output into the standard format
`InferenceData` using the Python library arviz [@kumarArviZUnifiedLibrary2019]
and saves it as a json file, along with some information for debugging.
This format allows for easy checking of MCMC diagnostics including divergent
transitions, effective sample size and the improved $\hat{R}$ statistic proposed
in @vehtariRankNormalizationFoldingLocalization2021.

## Solving the steady state problem

Maud's parameters are connected with measurable quantities via the steady state
equation \eqref{eq-steady}. Posterior sampling using adaptive Hamiltonian Monte
Carlo requires repeatedly solving equation \eqref{eq-steady} and finding the
gradients of its solution with respect to all model parameters. This must be
done numerically as analytic solutions are not available for realistic kinetic
models.

The speed at which this problem can be solved is tightly coupled with the
size and complexity of metabolic network that can practically be modelled. See
@timonenImportanceSamplingApproach2022a for more about considerations involved
in this kind of modelling.

To solve equation \eqref{eq-steady} and find its gradients, Maud uses
a hybrid method involving two numerical solvers from the SUNDIALS
suite [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via
their interface from Stan. The hybrid method follows that proposed by
@margossianComputingSteadyStates2018, and involves numerically evolving the ODE
system for a short period of time, then using the difference between the evolved
and starting concentrations as the target for a numerical algebra solver.

### 2023-10-31 paper/main.tex
Update paper

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, enhanced, boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\includegraphics{./figures/methionine-reactions.png}

@@

\includegraphics{./figures/missing.png}

@@

\includegraphics{./figures/decomposition.png}

>>>>

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

\includegraphics{../figures/methionine-reactions.png}

@@

\includegraphics{../figures/missing.png}

@@

\includegraphics{../figures/decomposition.png}

### 2023-10-31 paper/supplementary.qmd
Update paper

>>>>

# Maud's input format

Maud inputs are structured directories, somewhat inspired by the PEtab
format [@SchmiesterSch2021]. A Maud input directory must contain a toml
[@preston-wernertomandgedampradyunTOMLSpecification0rc2020] file called
`config.toml` which gives the input a name, configures how Maud will be run
and tells Maud where to find the other files, allowing these to have custom
names. It must also include a file containing a kinetic model definition,
a file specifying information about parameters and a file with information
experiments. The required structure of these files is documented at <https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html>. The input is
validated against a Pydantic [@pydanticdevelopersPydantic2022] data model which can be
inspected at <https://github.com/biosustain/Maud/tree/main/maud/data_model>.

We chose to implement a custom input format despite the existence of standard
formats in similar areas, including SBML [@keatingSBMLLevelExtensible2020]
and PEtab [@SchmiesterSch2021]. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features of
SBML and PEtab that make them structurally unsuitable in this context. Our
requirements for an input format included that it be mathematics-free, so that
all mathematical details are encapsulated in source code, and that it has a
detailed, verifiable structure. These requirements made toml more attractive
than SBML: toml is easier for humans to read and edit and can straightforwardly
be validated using tools like Pydantic. Further, an SBML representation of our
desired input would not contain differential equations.  It would therefore not
be interoperable with most SBML targeting software, which typically assumes that
differential equations are available and does not know about Maud’s structure.
